Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-08-31 paper.tex
discussion of discordances

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}

>>>>

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}
To assess robustness of the inferred models to analysis and reference population choices, 
Sections \ref{sec:validation_strategy} and \ref{sec:validation_results} present re-analyses changing the European and West African populations, as well as the recombination maps, 
filtering strategies and parameter optimization strategies. While we find some differences in the inferred parameters (see in particular \ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all reanalyses are quantitatively consistent.

### 2022-08-31 supplement.tex
discussion of discordances

\section{Model validation}

We took two approaches to validate demographic model inferences. First, we
explored the effects of using alternative populations, sample sets, and
recombination maps, and the effect of relaxing fixed parameters when fitting
the demographic models to diversity and LD statistics. We then use simulations
to test our inferred model predictions against other informative summary
statistics. Results are shown in the Supplementary Results
(Section~\ref{sec:supp-results}).

@@

\section{Supplementary results}\label{sec:supp-results}

\subsection{Model fit using alternative datasets}
\label{sec:supp-alt-data}

@@

\subsubsection{Replacing Thousand Genomes populations}

We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
and Western and Northern Europeans (CEU), respectively, and refit our inferred
demographic models to diversity and LD statistics computed from those
populations (along with Nama, Oromo/Amhara, and Gumuz). Because the MSL and YRI
share recent West African ancestry, and the GBR and CEU share recent European
ancestry, we expected most parameters in inferred models using these two
datasets to be largely similar. Indeed, all parameters inferred using the YRI
and CEU data were well within one standard error of the same parameters
inferred using the MSL and GBR data, across all tested models. The one
exception to this was the recent size of the YRI branch, inferred to be
somewhat smaller than the final size of the analogous MSL branch ($\approx
20,000$ vs. $\approx 30,000$, resp.). Other population sizes, migration rates,
and divergence times were highly consistent.

@@

In the demographic inference results presented in the main text, we did not
impose a minimum threshold on inferred Nama ancestry (as estimated using
AMDIXTURE). Because our proposed models allow for post-divergence gene flow as
well as recent admixture events, we can jointly learn deeper history and recent
admixture dynamics. This is in contrast to methods that do not account for
ongoing migration or recent admixture events, for which un-admixed genomes (if
they even exist) are needed for unbiased estimates of earlier history.
Although the methods used in this paper (\texttt{moments-LD}) are able to infer
early and recent history jointly, we tested the robustness of the inferred
early history to varying ancestry proportions in the Nama population. To do
this, we used ADMIXTURE to cluster inferred ancestry components, noting that
ADMIXTURE-inferred ancestries are coarse and can be misleading about true
underlying admixture proportions. Nonetheless, we included Nama individuals
that exceeded either 90\% ro 99\% ancestry primarily shared among Nama
individuals.

@@

\subsection{Mutation versus recombination rates}
\label{sec:supp-mutation}

@@

>>>>

\section{Model validation strategy}
\label{sec:validation_strategy}
We took two approaches to validate demographic model inferences. First, we
explored the effects of using alternative populations, sample sets, and
recombination maps, and the effect of relaxing fixed parameters when fitting
the demographic models to diversity and LD statistics. We then use simulations
to test our inferred model predictions against other informative summary
statistics. Results are shown in the Supplementary Results
(Section~\ref{sec:supp-results}).

@@

\section{Model validation results}\label{sec:supp-results}
\label{sec:validation_results}
\subsection{Model fit using alternative datasets}
\label{sec:supp-alt-data}

@@

\subsubsection{Replacing Thousand Genomes populations}
\label{sec:replacing}
We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
and Western and Northern Europeans (CEU), respectively, and refit our inferred
demographic models to diversity and LD statistics computed from those
populations (along with Nama, Oromo/Amhara, and Gumuz). Because the MSL and YRI
share recent West African ancestry, and the GBR and CEU share recent European
ancestry, we expected most parameters in inferred models using these two
datasets to be largely similar. Indeed, all parameters inferred using the YRI
and CEU data were well within one standard error of the same parameters
inferred using the MSL and GBR data, across all tested models. The one
exception to this was the recent size of the YRI branch, inferred to be
somewhat smaller than the final size of the analogous MSL branch ($\approx
20,000$ vs. $\approx 30,000$, resp.). Other population sizes, migration rates,
and divergence times were highly consistent.

@@

In the demographic inference results presented in the main text, we did not
impose a minimum threshold on inferred Nama ancestry (as estimated using
AMDIXTURE). Because our proposed models allow for post-divergence gene flow as
well as recent admixture events, we can jointly learn deeper history and recent
admixture dynamics. This is in contrast to methods that do not account for
ongoing migration or recent admixture events, for which un-admixed genomes (if
they even exist) are needed for unbiased estimates of earlier history.
Although the methods used in this paper (\texttt{moments-LD}) are able to infer
early and recent history jointly, we tested the robustness of the inferred
early history to varying ancestry proportions in the Nama population. To do
this, we used ADMIXTURE to cluster inferred ancestry components, noting that
ADMIXTURE-inferred ancestries are coarse and can be misleading about true
underlying admixture proportions. Nonetheless, we included Nama individuals
that exceeded either $90\%$ or $99\%$ ancestry primarily shared among Nama
individuals.

@@

@@

\section{Mutation versus recombination rates}
\label{sec:supp-mutation}
